---
title: "Week 2"
output: html_notebook
---

```{r setup, include=FALSE}
library(tidyverse)
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
```

# Introduction to random variables

In realty - only subset of population is available --\> use for interference

## Exercise random vars.

### Download data

```{r download, message=FALSE}
myData<-read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv")
glimpse(myData)

head(myData)
```

### Expercises

```{r}
meanWeight<-mean(myData$Bodyweight)
set.seed(1)
meanSubs<-mean(sample(myData$Bodyweight,5))
abs(meanWeight-meanSubs)

set.seed(5)
meanSubs<-mean(sample(myData$Bodyweight,5))
abs(meanWeight-meanSubs)
```

# Null distributions and p values

```{r}
femaleMiceControl<-read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv")

femaleMiceControl

# Simulate no treatment effect --> H0 == T
# run 1000 iterations
n<-10000
nulls<-vector(length = n)

for(i in 1:n){
  control<-sample(femaleMiceControl$Bodyweight,12)
  treatment<-sample(femaleMiceControl$Bodyweight,12)
  nulls[i]<-(mean(control)-mean(treatment))
}

# Show likelyhood of extreme differences under null hypothesis 
hist(nulls)
```

### count how often a vlue \> observed appears in Nulls

```{r}
# load data
read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv")->femaleMice

femaleMice%>%group_by(Diet)%>%
  summarise(mean(Bodyweight),.groups = "drop")->groupedMeans


obsDiff<-groupedMeans[2,2] - groupedMeans[1,2]

# Compute proortion of null distributed values that are more extreme than the observed once 

sum(nulls > obsDiff$`mean(Bodyweight)`) /n ->oneSided
sum(abs(nulls) > obsDiff$`mean(Bodyweight)`) /n ->twoSided

```

Probability that the outcome from the null distribution is more extreme than the observed value when H0 is true: For one sided: `r oneSided`\
Two sided: `r twoSided`

In the given example, the null distribution was constructed based on a complete dataset of the non-trated population. It is usually not possible to study the whole population --\> that is done in the next steps

## Null distribution exercise

```{r}
femaleMiceControl<-unlist(femaleMiceControl)
popMean<-mean(femaleMiceControl)
set.seed(1)
n=10000
splMeans<- vector("numeric",n)
for (i in 1:n) {
  sample(femaleMiceControl,5)%>%mean()->splMeans[i]
}


sum(abs(popMean-splMeans) > 1) /n
```

# Propability distributions

## How to summarise data

-   Histogram is a good summary

    -   helps to compute proportion of samples below / above / within certain value

-   Empirical cumulative distribution function

    -   f(x) for x == observed valuable and f(x) cumulative proportion (i.e. proportion of all values up to x)

    -   shows proportion of values less or equal to x

    -   f(x) can be computed by mean(sample \<= x)

        -   calculates sum of values in sample that are less or equal to x and divides by length of sample

## Exersice propability distribution

### load data

```{r}
library(gapminder)
data(gapminder)
head(gapminder)
```

### Proportion of countries in 1952 with lifeExpct \<=40

```{r}
gapminder%>%filter(year==1952)%>%
  summarise(mean(lifeExp<=40))
# same:
gapminder%>%filter(year==1952)%>%
  summarise(sum(lifeExp<=40)/nrow(.))

```

### builid a custom ecdf function

```{r}
gapminder%>%filter(year==1952)%>%select(lifeExp)->x

prop<-function(q){
  out<-mean(x <= q)
  return(out)
}

prop(40)
```

### Sequence over several qs

```{r}
qs<-seq(min(x),max(x),1)

sapply(qs, prop)->props
plot(qs,props)
```

### Use build in ecdf function to do the same

```{r}
plot(ecdf(x$lifeExp))

```

# Central limit theorem

Normal approximation --\> needs only mu (average) and sigma (standard deviation) to define distribution + integral can be used to estimate proportion of all values within an interval + QQ plots compute for each percentile the number of data that is expected to be below that data and plot explected vs. observed + Standard units + If data is aprox normal -\> can be converted into standard units by substracting mean and dividing by SD --\> normal distribution with mean = 0 and SD = 1 --\> Z score (standard normal distribution) Z score == amount of SDs far away from mean

## Normal distribution exercise

### take samples and assess distribution

```{r}
# make averages5
set.seed(1)
n <- 1000
averages5 <- vector("numeric",n)

for(i in 1:n){
  X <- sample(femaleMiceControl,5)
  averages5[i] <- mean(X)
}

# make averages50
set.seed(1)
n <- 1000
averages50 <- vector("numeric",n)
for(i in 1:n){
  X <- sample(femaleMiceControl,50)
  averages50[i] <- mean(X)
}

par(mfrow=c(1,2))
hist(averages5)
hist(averages50)
```

### Proportion of averages between 23 and 25

```{r}
avFilt<-averages50>=23 & averages50<=25
mean(avFilt)
```

### Generalised: find poroportion of observations between 23-25 with mu=23.9 and sigma =0.43 (like the data above...)

-   use pnorm function

    -   pnorm(x, mu, sigma)
    -   x= cutof

```{r}
pnorm(25,23.9,0.43)-pnorm(23,23.9,0.43)


```

## Populations, parameteres and sample estimates

-   if only samples from populations can be taken (not entire population) --\> statistical inference

    -   take random measurements from populations

Population average --\> mu of different groups are searched and compared

Sample averages are random variables

## Populations exercise

### get data

```{r}
micePheno<-read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv")%>%na.omit()

glimpse(micePheno)
head(micePheno)
micePheno%>%filter(Sex=="M",Diet=="chow")%>%select(Bodyweight)%>%
  unlist()->maleChow

micePheno%>%filter(Sex=="M",Diet=="hf")%>%select(Bodyweight)%>%
  unlist()->maleHf
```

### mean and population SD

```{r}
library(rafalib)
mean(maleChow)
popsd(maleChow)

mean(maleHf)
popsd(maleHf)

```

### sample means

```{r}
set.seed(1)
mean(sample(maleChow,25))->maleChowSampleMean
set.seed(1)
mean(sample(maleHf,25))->maleHfSampleMean
```

### abs difference between differences in sample and population means

```{r}
mean(maleChow) - mean(maleHf) ->malePopDif
maleChowSampleMean - maleHfSampleMean -> maleSampleDif

abs(malePopDif - maleSampleDif)
```

```{r}
micePheno%>%filter(Sex=="F",Diet=="chow")%>%select(Bodyweight)%>%
  unlist()->femaleChow

micePheno%>%filter(Sex=="F",Diet=="hf")%>%select(Bodyweight)%>%
  unlist()->femaleHf
```

### abs difference between differences in sample and population means

```{r}
set.seed(2)
mean(sample(femaleChow,25))->femaleChowSampleMean
set.seed(2)
mean(sample(femaleHf,25))->femaleHfSampleMean

mean(femaleChow) - mean(femaleHf) ->femalePopDif
femaleChowSampleMean - femaleHfSampleMean -> femaleSampleDif

abs(femalePopDif - femaleSampleDif)
```

Central limit theorem --\> sample average follows normal distribution with mean population average and $sigma = sigmaX/sqrt(N)$ + SigmaX --\> population standard deviation (typical deviation from mean) + Sample average --\> random variable + repeated samples --\> sample means would follow the aforementioned normal distribution $N(mu_x,\frac{sigma_x}{\sqrt{N}})$  

How to use this in practice? 
Differences betwen two sample means: 

$N(0,1) = \frac{Y - X}{\sqrt{{sigma_x^2}/{N} + sigma_y^2/N}}$ if H0 == TRUE
 --> with increasing N, SD of final distribution decreases, histogram gets "narrower". 
+ Problem in real life: Population SD is not known --> t.test is needed


## Exercise CLt
If list of numbers is normally distributed, what proportion of these numbers are within 1 SD of the average?
```{r}
pnorm(3) - pnorm(-3)  

```

### proportin of male control mice within 1Sd from average weight
```{r}
maleChowMean<-mean(maleChow)
maleChowSD<-popsd(maleChow)

1- (mean(maleChow >=maleChowMean+maleChowSD) + mean(maleChow <=maleChowMean-maleChowSD))

# W/in 2sd
1- (mean(maleChow >=maleChowMean+2*maleChowSD) + mean(maleChow <=maleChowMean-2*maleChowSD))

# W/in 3sd
1- (mean(maleChow >=maleChowMean+3*maleChowSD) + mean(maleChow <=maleChowMean-3*maleChowSD))

```
### test CLT
```{r}
set.seed(1)
avgs<-replicate(10000, mean(sample(maleChow,25)))
mypar(1,2)
hist(avgs)
qqnorm(avgs)
qqline(avgs)
mean(avgs)
popsd(avgs)

```

