---
title: "Week 1"
author: "Jonathan Speh"
date: "1/23/2021"
output: html_document
---

```{r}
library(rafalib)
library(readr)
library(dplyr)
library(ggplot2)
```


```{r}
myVec<-c(2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23 )
mean(myVec)
```
# Use looop to determine valeu of sum (i 1:25) i²

+ why would I
```{r}
sum(c(1:25)^2)
```
# some cars stuff
```{r}
data(cars)
class(cars)
nrow(cars)
colnames(cars)
mean(cars[,2])

which(cars$dist==85)


```
# Female mice dataset
```{r}
femaleMice<-read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv")

head(femaleMice)
```
```{r}
dim(femaleMice)
```

## average weigh on high fat diest

```{r}
# Thats how they want me to do it...
# femaleMice[which(femaleMice$Diet=="hf"),]
# femaleMice[femaleMice$Diet=="hf",]

femaleMice%>%group_by(Diet)%>%summarise(meanWeight=mean(Bodyweight),.groups="drop")
```
## sample funciton

```{r}
set.seed(1)
femaleMice[sample(13:24,1),]

```

# sleep dataset
```{r}
sleepDat<-read_csv("https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv")

class(sleepDat)
head(sleepDat)
```
## explore
```{r}
sleepDat%>%count(order,sort=TRUE)

sleepDat%>%filter(order=="Primates")%>%select(sleep_total)%>%class()
sleepDat%>%group_by(order)%>%summarise(meanSleep=mean(sleep_total))
```


# EDA
## QQ plot

+ that's tha part that didn't work because the skew data is a mess
```{r}
skewData<-base::load("~/Downloads/skew.RData")


dim(skewData)
```

# Not normally distributed data

```{r}
data("InsectSprays")
head(InsectSprays)
```
##  boxplots
+ which has then  lowees median
+ Boxplot components. 
  + Median
  + 25th and 75 quantile --> box = 25% of data
  + range
  + outliers
```{r}
boxplot(count ~ spray, data = InsectSprays)
```


```{r}
data(nym.2002, package = "UsingR")
head(nym.2002)
```


```{r}
nym.2002%>%
  ggplot(aes(x=time))+
  geom_histogram(alpha=0.7,colour="black")+
  facet_wrap(~gender)
```

```{r}
nym.2002%>%
  ggplot(aes(x=gender, y=time))+
  geom_violin(draw_quantiles = c(0.25,0.5,0.75))
  
nym.2002%>%group_by(gender)%>%
  summarise(mean=mean(time),median=median(time))
  
```

